---
title: "PA1_Activity_Monitoring.Rmd"
author: "Lynn Scott"
date: "July 11, 2016"
output: html_document
---


## Load and pre-process data.

```{r echo=TRUE}
setwd("/Users/rs186089/coursera/repro-research/RepData_PeerAssessment1")
unzip("activity.zip")
activity.data <- read.table("activity.csv", header=TRUE, sep=",")

str(activity.data)
```

## What is the mean total number of steps taken per day?

```{r echo=TRUE}

library(dplyr)

total.steps.per.day = aggregate(activity.data$steps, list(date = activity.data$date), sum, na.rm = TRUE)
colnames(total.steps.per.day)[2] = "total.steps"

head(total.steps.per.day)
```

Below is the distribution of the total number of steps taken per day (missing values are ignored):

```{r echo=TRUE}
hist(total.steps.per.day$total.steps, 
     main = "Distribution of Total Number of Steps per Day", 
     xlab = "Total Number of Steps per Day", 
     ylab = "Frequency (Number of Days)", 
     breaks=20, col = "blue")
```

### Mean of steps taken per day:
```{r echo=TRUE}
mean(total.steps.per.day$total.steps, na.rm=TRUE)
```
### Median of steps taken per day:
```{r echo=TRUE}
median(total.steps.per.day$total.steps, na.rm=TRUE)
```


## What is the average daily activity pattern?

Calcuate the average number of steps per 5-minute interval over the 2 month period:

```{r echo=TRUE}

avg.steps.per.interval = aggregate(activity.data$steps, list(interval = activity.data$interval), mean, na.rm = TRUE)

colnames(avg.steps.per.interval)[2] = "avg.steps"

head(avg.steps.per.interval)

```
The following graph represents the average daily activity pattern. It is a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):

```{r echo=TRUE}
plot(strptime(sprintf("%04d", avg.steps.per.interval$interval), format="%H%M"),
     avg.steps.per.interval$avg.steps, type = "l", 
     main = "Average Daily Activity", 
     xlab = "Time of Day (HH:MM)", 
     ylab = "Average Number of Steps")


```

By deriving the maximum of average steps from avg.setsp.per.interval, the interval the containst the maximum number of steps is show:

```{r echo=TRUE}
  max.interval = filter(avg.steps.per.interval, avg.steps==max(avg.steps))
```

So the 5 minute interval that contains the maximum number of steps is: `r max.interval[1]`.

## Inputing missing values:

Total number of missing values:

```{r echo=TRUE}
  sum(is.na(activity.data))
```

New data set with missing values is replaced with the mean for each interval of the original data set.

```{r echo=TRUE}

activity.data.revised = inner_join(activity.data, avg.steps.per.interval, by="interval") %>%
                    mutate(steps=ifelse(is.na(steps), avg.steps, steps)) %>%
                    select(date, interval, steps)

head(activity.data.revised)
```

Below is the distribution of the total number of steps taken per day (NAs have been normalized to contain the mean of the interval):

```{r echo=TRUE}

total.steps.per.day = aggregate(activity.data.revised$steps, list(date = activity.data.revised$date), sum, na.rm = TRUE)
colnames(total.steps.per.day)[2] = "total.steps"

hist(total.steps.per.day$total.steps, 
     main = "Distribution of Total Number of Steps per Day (revised)", 
     xlab = "Total Number of Steps per Day", 
     ylab = "Frequency (Number of Days)", 
     breaks=20, col = "blue")
```

With the data normalized to revise missing intervals the mean does not change:
```{r echo=TRUE}
mean(total.steps.per.day$total.steps, na.rm=TRUE)
```
but the median has changed to equal the mean.
```{r echo=TRUE}
median(total.steps.per.day$total.steps, na.rm=TRUE)
```

### Are there differences in activity patterns between weekdays and weekends?

First make factors for the the days of the week (i.e. weekday/weekend).

```{r echo=TRUE}

weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
activity.data.revised$day.of.week = as.factor(ifelse(is.element(weekdays(as.Date(activity.data.revised$date)),weekdays), "Weekday", "Weekend"))
                                   
head(activity.data.revised)
```

Calculate the average number of steps per calandar day:
```{r echo=TRUE}
activity.pattern = activity.data.revised %>%
                    group_by(day.of.week, interval) %>%
                    summarize(avg.steps=mean(steps))

head(activity.pattern)
```


Plot the average daily activity.
```{r echo=TRUE}

library(ggplot2)
library(scales)

ggplot(activity.pattern, aes(interval, avg.steps, group=1)) +
    geom_line() +
    facet_grid(day.of.week~., scale='free_y') +
    xlab('Time of Day') +
    ylab('Average Number of Steps') +
    ggtitle("Average Daily Activity\nWeekday vs. Weekend")

```

The following conclusions can be drawn from the plot above:

* People are active in the early morning weekeday then the weekeend days
* Weekday ctivity is quite intense around 8:30AM
* Weekend activity is more spread out during the day